---
title: "3. Introduction to hypothesis testing via binomial tests"
bibliography: ../references.bib
editor: 
  markdown: 
    wrap: 72
---

<!-- COMMENT NOT SHOW IN ANY OUTPUT: Code chunk below sets overall defaults for .qmd file; these inlcude showing output by default and looking for files relative to .Rpoj file, not .qmd file, which makes putting filesin different folders easier  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

Remember you should

-   add code chunks by clicking the *Insert Chunk* button on the toolbar
    or by pressing *Ctrl+Alt+I* to answer the questions!
-   **knit** your file to produce a markdown version that you can see!
-   save your work often
    -   **commit** it via git!
    -   **push** updates to github

## Overview

This practice reviews the [Hypothesis testing starting with binomial
tests lecture](../chapters/Binomial.qmd).

## Hypothesis Testing and the Binomial Distribution

### Example

Using the bat paper from class (Geipel et al. 2021), let's consider how
to analyze data showing all 10 bats chose the walking over the
motionless model.

```{r}
binom.test(10,10)
```

We use the binom.test function. We only need arguments for \# of
succeses and \# of trials. By default it runs a 2-sided test against a
null hypothesis value of p = .5. You can see how to update thee options
by looking at the help file.

```{r, eval=F}
?binom.test
```

Note the confidence interval is assymetric since its estimated to be 1!
We can see other options using the binom.confint function from the
*binom* package.

```{r}
library(binom)
binom.confint(10,10)
```

All of these correct for the fact that most intervals use a normal
approximation, which as you remember from our earlier discussions is not
good when sample sizes are small and/or the p parameter is extreme
(close to 0 or 1).

## Practice!

Make sure you are comfortable with null and alternative hypotheses for
all examples.

### 1

Are people eared (do they prefer one ear or another)? Of 25 people
observed while in conversation in a nightclub, 19 turned their right ear
to the speaker and 6 turn their left ear to the speaker. How strong is
the evidence for eared-ness given this data (adapted from Analysis of
Biological Data)? \* state a null and alternative hypothesis \*
calculate a test statistic (signal) for this data \* Make sure you
understand how to construct a null distribution + using
sampling/simulation (code or written explanation) + by using an
appropriate distribution (code or written explanation) \* Calculate and
compare p-values obtained using + simulation (calculation won’t be
required on test, but make sure you understand!) (code or written
explanation) + equations for binomial distribution (code or written
explanation) + R functions (required)(code) \* Calculate a 95%
confidence interval for the proportion of people who are right-eared \*
How do your 95% confidence interval and hypothesis test compare?

**Hypotheses**

H~0~: People are not eared; the proportion of people turning their ear
to the right is equal to 0.5 (p=0.5)

H~A~: People are eared; the proportion of people turning their ear right
is different than 0.5 (p does not equal 0.5)

**Test Statistic**

Signal = observed proportion of right ear - expected portion

Signal = 19/25 - 1/2 = 0.26

**Null Distribution**

To construct a null distribution using sampling or simulation, I would
simulate data under the assumption that the null hypothesis is true.
This means generating data that represents what you would expect to
observe if there were no effect or difference. The null hypothesis
assumes no preference, meaning the probability of someone turning their
right ear to the speaker is 0.5, and the probability of turning their
left ear to the speaker is also 0.5.

```{r}

n <- 25  
p_null <- 0.5 

num_simulations <- 10000

null_distribution <- numeric(num_simulations)

for (i in 1:num_simulations) {
 
  simulated_data <- sample(c("right", "left"), n, replace = TRUE, prob = c(p_null, 1 - p_null))
  
  observed_proportion <- sum(simulated_data == "right") / n
  
  null_distribution[i] <- observed_proportion
}

```

```{r}
head(null_distribution[i])
```

```{r}
library(ggplot2)

ggplot() +
  geom_histogram(data = data.frame(observed_proportion = null_distribution),
                 aes(x = observed_proportion),
                 bins = 
```

```{r}

n <- 25
observed_right <- 19


p_null <- 0.5


B <- 10000  
null_distribution <- rbinom(B, n, p_null) / n


null_mean <- mean(null_distribution)
signal <- observed_right / n - null_mean

signal
```

**p-values by simulation and binomial**

```{r}

n <- 25
observed_right <- 19

p_null <- 0.5

B <- 10000  
null_distribution <- rbinom(B, n, p_null) / n

null_mean <- mean(null_distribution)
signal <- observed_right / n - null_mean

p_value_simulation <- 2 * mean(abs(null_distribution - null_mean) >= abs(signal))

p_value_simulation
```

```{r}

n <- 25
observed_right <- 19

p_null <- 0.5

p_value_binomial <- pbinom(observed_right - 1, n, p_null) + (1 - pbinom(observed_right, n, p_null))

p_value_binomial <- 2 * p_value_binomial

p_value_binomial
```

The p-values are very different, as the binomial one is a lot greater
than the simulated one.

```{r}

p_value <- 2 * pnorm(-abs(signal))

p_value
```

```{r}

n <- 25
observed_right <- 19

p_null <- 0.5

B <- 10000  
null_distribution <- rbinom(B, n, p_null) / n

```

```{r}
binom.test(x=19, n=25, p = 0.5,
           alternative = c("two.sided", "less", "greater"),
           conf.level = 0.95)
```

The confidence interval provides a range of values for the true
proportion of people who turn their right ear to the speaker, while the
hypothesis test determines whether there is significant evidence to
reject the null hypothesis that the proportion is equal to 0.5. If the
confidence interval includes 0.5, it suggests that the observed
proportion is not significantly different from 0.5 at the 5%
significance level. If the confidence interval does not include 0.5 and
the p-value is less than 0.05, it suggests that there is significant
evidence to reject the null hypothesis in favor of the alternative
hypothesis.

```{r}

z <- qnorm(0.975)
SE <- sqrt(p_null * (1 - p_null) / n)
lower <- observed_right / n - z * SE
upper <- observed_right / n + z * SE
```

```{r}

n <- 25
observed_right <- 19

p_null <- 0.5

B <- 10000  
null_distribution <- rbinom(B, n, p_null) / n

null_distribution
```

### 2

A professor lets his dog take every multiple-choice test to see how it
compares to his students (I know someone who did this). Unfortunately,
the professor believes undergraduates in the class tricked him by
helping the dog do better on a test. It’s a 100 question test, and every
questions has 4 answer choices. For the last test, the dog picked 33
questions correctly. How likely is this to happen, and is there evidence
the students helped the dog?

**MAKE SURE TO THINK ABOUT YOUR TEST OPTIONS**
